Analysis of competitive surfing tournaments with generalized Bradley-Terry likelihoods

### 2022-09-25 inst/surfing_jsa_rebuttal_minorcomments.tex

\documentclass[12pt]{article}
\usepackage{xcolor}
\begin{document}

\section*{MS596 ``Analysis of competitive surfing tournaments with generalized Bradley-Terry likelihoods": response to minor comments}

Below, we reproduce the reviewers' minor comments; our replies to
the issues are in \textcolor{blue}{blue}.  We have indicated changes
to the manuscript where appropriate.

@@

Short story: each of the  comments is perfectly reasonable and we have 
changed the manuscript as suggested.

Overview I am pleased to see the authors have addressed the issues
identified in the first review of the paper, or in one or two cases,
have provided reasonable justification for their approach. The changes
to the general structure have improved the paper’s overall readability
and flow, and I commend the inclusion of more explicit derivations and
references to numerical quantities. As such, I am satisfied with the
revisions the authors have made and I am happy to recommend the paper
be accepted, conditional on some minor revisions, noted below:

Contextual changes and clarifications

Page 2: The first reference to the ``World Championship Tour'' would
benefit from including a one-off abbreviation (WCT), which is referred
to throughout the data set.

\textcolor{blue}{Done; WSL used throughout for ``World Surf League''
  and minor rewording ensures that the phrase ``World Championship
  Tour'' does not appear}

Page 4: It is mentioned in the abstract that the analysis is performed on data from the 2019 WSL,
however a reference in the main text at the top of Section 5 would help clarify this.
Page 4: The authors mention the surfer Vries – who does not appear to be a wildcard – whose
maximum likelihood strength is zero. Why is this? My guess is that Vries has never won a competitive
surfing heat (and therefore there is no information on his true, underlying ability), however a brief
clarification here would be useful.
Page 6 / Figure 1: I’m pleased to see the authors replace the old pie chart in Figure 1 with a dot chart
– this is a much clearer visualization of the results. On Page 6 the authors state: “Figure 1 shows the
maximum likelihood estimate for the 23 competitors’ strengths, which appears to show a wide range
from Florence at about 17.5% down to Morais at about 1%”. What exactly do the quantities 17.5%
and 1% refer to/how are they computed? A quick clarification of these quantities would help explain
the practical implications of the findings.
Page 7/Figure 2: The authors suggest there is a loose positive correlation between maximum likelihood
strength and total points, which is undoubtedly true. However, the x-axis on Figure 2 itself has been
distorted, which makes the trend appear more linear than it really is. I’d prefer to see the x-axis
undistorted, unless the authors are explicitly trying to make a point about the trend being log-linear,
in which case I’d like to see this briefly discussed.
Page 7: The authors suggest we may reject the hypothesis that the strength of the entity exceeds
about 0.025. Does this assertation come with a supporting p-value? It is unclear from Figure 4 how
this conclusion is reached.

Grammatical changes
Page 3: Change “Further, we use a generalization due to (Luce, 1959)...” to “Further, we use a
generalization due to Luce (1959)...”.
Page 7: Change “..., but in addition we may reject the the proposal that...” to “..., but in addition we
may reject the proposal that...”.

### 2022-09-25 inst/surfing_jsa.tex
fixes #167

\citet{farley2013} analyse surfing results using
one-way ANOVA techniques and report comparisons between the top 10 and
bottom 10 surfers in the 2013 World Championship Tour and conclude
that the top-ranked athletes are more consistent than the lower-ranked
group.  They report that ``to date, only a limited number of research
studies have reported on the results of competitive surfing" [p44],
and the analysis presented here fills some of this gap.

@@

and this would be a forward ranking Plackett-Luce model in the
terminology of \citet{johnson2020}.  We now use a technique due to
\citet{hankin2010,hankin2017} and introduce fictional (reified)
entities whose nonzero Bradley-Terry strength helps certain
competitors or sets of competitors under certain conditions.  The
original example was the home-ground advantage in football.  If
players (teams) $1,2$ with strengths $p_1,p_2$ compete, and if our
observation were $a$ home wins and $b$ away wins for team $1$, and $c$
home wins and $d$ away wins for team~$2$, then a suitable likelihood
function would be

@@

Figure~\ref{dotchartmax} shows the maximum likelihood estimate for the
23 competitors' strengths, which appears to show a wide range from
Florence at about 17.5\% down to Morais at about 1\%.  The {\tt
  hyper2} software~\citep{hankin2017} may be used to formally assess
one plausible null: that the competitors all have equal skill, that is
$H_0\colon p_\mathrm{Andino} = p_\mathrm{Bourez}=\ldots=
p_\mathrm{Wright}=\frac{1}{23}$, and any differences in placings are
due to random variation.  This test is performed by function {\tt
  equalp.test()} in the package which uses the method of support
\citep{edwards1992}.  The test rejects the null (in favour of the
evaluate) at $p=4\times 10^{-5}$.  There is thus strong evidence to
suggest that the competitors are indeed of different skills and the
result actually reflects genuine differences in abilities, rather than
the result of random chance.  Alternatively, we may view the dataset
as being sufficiently informative to reveal actual differences in
surfing ability.  Observe that the competitors' points (that is, the
final column of table~\ref{resultstable}) cannot be used for this type
of probabilistic analysis: the points awarded are arbitrary and
changing the points system would change the points scored but not
affect the competitors' behaviour.

@@

In surfing lore, the ``Brazilian storm'' refers to the emergence of
Brazil as a dominant nation of elite surfers.  The term was first used
to describe the fact that six out of eight quarterfinalists at the 2011 Nike
Lowers Pro were Brazilian; it rose in public consciousness following
the victories of Medina and de Souza in the World Surf League
championships in 2014 and 2015 respectively.  The storm shows no signs
of abating, with commentators such as Douglas opining that this
generation ``represents a seismic shift for Brazilian competitors onto
an entirely new plane'' \citep{douglas2020}.

@@

We analysed results from the 2019 World Surf League using reified
Bradley-Terry likelihoods to assess a number of hypotheses of interest
to the surfing community.  Reified Bradley-Terry allows the use of
formal statistical methods in ways that analysing points scored does
not.  For example, we can be confident that the results reflect
differences in skill (rather than random variation); there is no
statistical difference in strength between the top two officially
ranked competitors, Medina and Ferreira; but we may reject the
hypothesis that Medina and Wilson are equal in strength.

@@

\begin{table}[h]
\begin{tabular}{llllllllllllll}
competitor  & nationality&WCT01  &WCT02  &WCT03  & \ldots  &WCT10&WCT11&points\\ \hline
 Ferreira   &   BRA      &   1   &   5   &   17  & \ldots  &   1   &   1   &   59740\\
 Medina     &   BRA      &   5   &   5   &   17  & \ldots  &   9   &   2   &   56475\\
 Smith 	    &   ZAF      &   3   &   3   &   17  & \ldots  &   2   &   17  &   49985\\
 Toledo     &   BRA      &   9   &   2   &   5   & \ldots  &   5   &   17  &   49145\\
 Andino     &   USA      &   2   &   17  &   5   & \dots   &   5   &   9   &   46655\\
 $\vdots$   &            &   &   &       &       & $\vdots$&       &       &
\end{tabular}
\caption{Conventional league table for the top five competitors in WCT
  2019 tour. \label{resultstable} Thus at WCT01 (Coolangatta, QLD) we see
  Ferreira coming first, Andino second and Smith third}
\end{table}

>>>>

\citet{farley2013} analyse surfing results using
one-way ANOVA techniques and report comparisons between the top 10 and
bottom 10 surfers in the 2013 WSL tour and conclude
that the top-ranked athletes are more consistent than the lower-ranked
group.  They report that ``to date, only a limited number of research
studies have reported on the results of competitive surfing" [p44],
and the analysis presented here fills some of this gap.

@@

and this would be a forward ranking Plackett-Luce model in the
terminology of \citet{johnson2020}.  We now use a technique due to
\citet{hankin2010,hankin2017_nomarkup} and introduce fictional (reified)
entities whose nonzero Bradley-Terry strength helps certain
competitors or sets of competitors under certain conditions.  The
original example was the home-ground advantage in football.  If
players (teams) $1,2$ with strengths $p_1,p_2$ compete, and if our
observation were $a$ home wins and $b$ away wins for team $1$, and $c$
home wins and $d$ away wins for team~$2$, then a suitable likelihood
function would be

@@

Figure~\ref{dotchartmax} shows the maximum likelihood estimate for the
23 competitors' strengths, which appears to show a wide range from
Florence at about 17.5\% down to Morais at about 1\%.  The {\tt
  hyper2} software~\citep{hankin2017_nomarkup} may be used to formally assess
one plausible null: that the competitors all have equal skill, that is
$H_0\colon p_\mathrm{Andino} = p_\mathrm{Bourez}=\ldots=
p_\mathrm{Wright}=\frac{1}{23}$, and any differences in placings are
due to random variation.  This test is performed by function {\tt
  equalp.test()} in the package which uses the method of support
\citep{edwards1992}.  The test rejects the null (in favour of the
evaluate) at $p=4\times 10^{-5}$.  There is thus strong evidence to
suggest that the competitors are indeed of different skills and the
result actually reflects genuine differences in abilities, rather than
the result of random chance.  Alternatively, we may view the dataset
as being sufficiently informative to reveal actual differences in
surfing ability.  Observe that the competitors' points (that is, the
final column of table~\ref{resultstable}) cannot be used for this type
of probabilistic analysis: the points awarded are arbitrary and
changing the points system would change the points scored but not
affect the competitors' behaviour.

@@

In surfing lore, the ``Brazilian storm'' refers to the emergence of
Brazil as a dominant nation of elite surfers.  The term was first used
to describe the fact that six out of eight quarterfinalists at the 2011 Nike
Lowers Pro were Brazilian; it rose in public consciousness following
the victories of Medina and de Souza in the WSL
championships in 2014 and 2015 respectively.  The storm shows no signs
of abating, with commentators such as Douglas opining that this
generation ``represents a seismic shift for Brazilian competitors onto
an entirely new plane'' \citep{douglas2020}.

@@

We analysed results from the 2019 WSL using reified
Bradley-Terry likelihoods to assess a number of hypotheses of interest
to the surfing community.  Reified Bradley-Terry allows the use of
formal statistical methods in ways that analysing points scored does
not.  For example, we can be confident that the results reflect
differences in skill (rather than random variation); there is no
statistical difference in strength between the top two officially
ranked competitors, Medina and Ferreira; but we may reject the
hypothesis that Medina and Wilson are equal in strength.

@@

\begin{table}[h]
\begin{tabular}{llllllllllllll}
competitor  & nationality&WCT01  &WCT02  &WCT03  & \ldots  &WCT10&WCT11&points\\ \hline
 Ferreira   &   BRA      &   1   &   5   &   17  & \ldots  &   1   &   1   &   59740\\
 Medina     &   BRA      &   5   &   5   &   17  & \ldots  &   9   &   2   &   56475\\
 Smith 	    &   ZAF      &   3   &   3   &   17  & \ldots  &   2   &   17  &   49985\\
 Toledo     &   BRA      &   9   &   2   &   5   & \ldots  &   5   &   17  &   49145\\
 Andino     &   USA      &   2   &   17  &   5   & \dots   &   5   &   9   &   46655\\
 $\vdots$   &            &   &   &       &       & $\vdots$&       &       &
\end{tabular}
\caption{Conventional league table for the top five competitors in WSL
  2019 tour. \label{resultstable} Thus at WCT01 (Coolangatta, QLD) we see
  Ferreira coming first, Andino second and Smith third}
\end{table}
